Molecular dynamics simulation of synchronization in driven particles

### 2021-01-19 manuscript/manuscript_ajp_2020.tex
worked on constants and exercises

We confine the particles using a position dependent 
potential energy function, called a landscape or substrate.
 %The landscape potential energy $U(\vec{r})$ is static 
 %with fixed minima and maxima
 %that are periodic or quasi-periodic,
 %as described in Sec.~\ref{sec:results}.
% 
The landscape is modulated in the y-direction
with the periodic function 
%% %that used in Sec.~\ref{sec:one}
% where 
 \begin{equation}
   U(y) = U_0 \cos{(N_p \pi y / L)}
     \label{eq:ysubstrate}
\end{equation}
 where $N_p$ are the number of periods,
 % $L$ is the period length,
 and $U_0$ is an adjustable parameter
 to set the depth of the minima
 with simulation units of energy $E_0 = 1$. 
 %We define the substrate strength to be
 %the maximum force the substrate can exert on a particle 
 %$A_p = 2\pi U_0/L$.
 %
 We plot this function in 
 Fig.~\ref{fig:landscape0}
 for $N_p = 3$. %to illustrate the features of
 In Fig.~\ref{fig:landscape0}(a) we show 
 the $x-y$ plane with a contour plot of $U(y)$ %Eq.~\ref{eq:ysubstrate}
 to illustrate
 the 2D potential energy landscape,
 where 
 the maxima are colored red and the minima colored blue.
 The confining force on a particle $i$
 %arises from a potential function 
 %creating a landscape of potential minima and maxima
 %that modify the force on a particle as a function of position
 is calculated as 
 $\vec{F}^{l}_i(\vec{r}_i) = -\nabla U_l(\vec{r}_i)$.
 In Fig.~\ref{fig:landscape0}(b) we plot the function
 $U(y)$ to illustrate how the magnitude
 $\vec{F}^{l}_i$ is calculated from particle position $y_i$.  
 %in the $y-$direction.

@@

\begin{table}[h!]
\centering
\caption{Simulation units and approximate values
  in experiment ~\cite{Juniper2015}.
{\it might be more appropriate to match Juniper parameters more precisely given our focus...}}
\begin{ruledtabular}
\begin{tabular}{c c c } %p{5cm}}
Quantity & Simulation Unit & Experimental range\\
\hline
length &  $a_0 = 1$ & $ a_0 \sim 1.5 \mu m$\\
energy & $E_0 = 1$ & $ < E_0 <$ \\
%electric potential & $V(r_{ij}) = E_0/r_{ij}$ \\
%energy & $E_0 = q^2{Z^*}^2/4\pi \epsilon \epsilon_0 a_0$ \\
%dimensionless interaction strength & $q$ \\
%effective colloidal charge & $Z^*$ \\
%solvent dielectric constant & $\epsilon \epsilon_0$\\
force & $F_0 = E_0 / a_0$ & $<F_0<$\\
velocity &  $ a_0 / \tau $ &  $ < v < $\\
viscosity coefficient & $\eta = F_0 \tau / a_0 $ & $ <\eta<$\\
%juniper uses water–ethanol mixture
time &  $ \tau = \eta a_0 / F_0 $ & $ < \tau <$\\

@@

A single particle %on this landscape
responds to the applied driving force
by moving across the landscape %maxima
synchronized with 
the period of $F^d(t)$.
%applied driving force.
%along the y-direction
%({\it redundant, just refer back to eq. in Model?})
%and $V_{0y}=0.2$.
%This is illustrated in Fig.~\ref{fig:1}(a) where
%the red (blue) regions show local maxima (minima).
%
The numerical implementation of the landscape 
is calculated with Eq.~\ref{eq:ysubstrate} as 
\begin{equation}
  \label{eq:force}
  F^l_y(y) = -A_{p} \sin{(N_p \pi y / L)} 
\end{equation}
where the force is scaled with parameter $A_{p}$.
When $F^d(t) > A_p$, a particle can 
overcome the barrier height of the landscape,
and 
the particle hops between minima in the energy landscape.
In this section, we fix the landscape parameters
to $A_{p} = 0.1$ force units
with $N_p=20$ minima in the landscape
corresponding to a spatial period $\lambda = 1.825 a_0$.

%In Fig.~\ref{fig:0} 
The driven particle adjusts
its motion to traverse
the periodic landscape with the same rhythm as 
the applied time-dependent force $F^d(t)$.
%
In Fig.~\ref{fig:0}(a)
we plot $F^d(t)$ %defined in Eq.~\ref{eq:drive}.
as a function of time with
%The time independent
constants $F^{dc}=0.1$, 
$F^{ac}=0.05$ and $f=0.01$ cycles per time unit $\tau$.
The temporal period of the driving force is
$T = 1/f = 100 \tau$.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%RAMP
%To avoid transient oscillations at early times, the dc driving force is initially zero and slowly increased %$F^{dc}$. at a rate of $\Delta F^{dc} = 0.01$ every $\Delta t = 4$ time units  to the constant value $F^{dc}=0.1$. In Fig.~\ref{fig:0}(a) we plot $F^{dc}$ in blue to show the increase in  $F^{dc}$ for the first $40$ time units In orange we show  the combined ac and dc drive $F^d(t)$ when the magnitude of the ac applied force is initially zero.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%since $\sin{(0)}=0$.
We mark the maxima of $F^{ac}(t)$
with the
vertical dashed lines %in panels (a) and (b)
-- i.e. $\sin{(\omega t)} = 1$.
%
%
%with the initial driving force at is minimum value 
%$F^d(0) = F^{dc}$. 
%
In Fig.~\ref{fig:0}(b) 
we show the $y-$position of the particle
as a function of time.
We 
normalize $y$ by $\lambda$ %, the spatial period of the potential
to show  particle motion between substrate minima.  
The initial particle position is $y = 0 $. %1 a_0$,
%where the $F^l$ is negative.
%JUST BECAUSE
%causing the particle to move to
%a landscape minima at the start of the simulation.
As the driving force increases,
%the effects of the landscape become less obvious,
the motion of the particle
synchronizes with the driving period $T$.
We measure 
average particle velocity $\langle{v}_y\rangle$
as the average displacement %$\Delta y = y(t_0+T) - y(t_0)$
over the period of the driving force.
The particle moves
in the positive y-direction
through $\Delta y = 2 \lambda$ over the time period $T$,
with 
the average velocity 
$\langle {v}_y \rangle= 2 \lambda f$. % where $T = 1/f$.
The horizontal dashed lines
coincide with %the condition the is
minimum substrate force  $F^l = -A_p$,  
i.e. $\sin{(\pi y / \lambda)} = 1$.
%or $\pi y / \lambda = m \pi/2$, where $m = 1, 5, 7$, {\it etc.}
%Once the particle motion
%coincides with the applied drive,
%indicating synchronization,
The particle synchronizes its motion such that the 
driving force is maximum when the landscape 
force is minimum,
as indicated with %the coincidence of 
with the intersection of
horizontal and vertical lines
and the particle position $y$.
The slope of $y/\lambda$
is proportional to the net force on the particle.
This can be seen in 
Fig.~\ref{fig:0} %we demonstrate this relationship,
where local extrema appear in $y/\lambda$
when $F_{D}(t) \approx 0 $.

@@

%the average velocity of the particle
%is constant 
%despite the overall magnitude
%of $F^{dc}$.
Each step represents a different pattern of hops
between substrate minima
performed by the particle
due to the landscape confinement.  
In Fig.~\ref{fig:1} at %this occurs at 
low $F^{dc}$ the average velocity $\langle v_y \rangle$ is zero.
Since 
$A_p$ is large compared to the extrema of $F^{d}(t)$,
the particle oscillates back and forth
in a single minima with no net velocity.
At higher $F^{dc}$ the particle velocity 
$\langle v_{y} \rangle$ increases in steps of uniform height,
$\langle v_{y} \rangle = n \lambda f$,
where $n$ is an integer.
The step width is non-linear 
and can have a variety of
interesting patterns \cite{Bak1986}.
%$\lambda = L/N_p = 36.5/20 = 1.825$ is the spatial period, or wavelength
%of the landscape,
%and $f = 0.01$ cycles per time unit.

@@

We 
show the dynamics 
in supplementary materials \cite{supp1}.

@@

[Insert Phase Plots section/figure?]

@@

  This model should be familiar to readers
  of a standard classical mechanics text, such as Ref.~\cite{Taylor2005}
  when studying 
  models Millikan's oil drops.
  [{\it could certainly do more with this question to make it more computational.  For instance, plot $R$ as a function of any of the parameters $v$, $D$ and $\rho$, include a temperature dependent $\eta(T)$ in the Brownian system.  Suggest students model something big and include inertia and quadratic drag, but then we're fairly off topic...}]

@@

\item{\bf Equation of Motion}

  Newton's second law states that
  the acceleration of a particle $i$
  is proportional to 
  the sum of forces on a particle 
  \begin{equation}
  m_i \vec{a}_i = \sum \vec{F}_i
  \label{eq:n2l}
  \end{equation}
  where the constant of proportionality is the
  inertial mass $m_i$
  The addition of a dissipative force in a dynamical model
  on a colloid is typically modeled
  with a drag force proportional to the particle's velocity
  in the opposite direction of motion 
  $\vec{F}^{drag} = - \eta \vec{v}_i$
  where $\eta$ is the drag coefficient.
  The ratio of $m/\eta$ has units of time, 
  %This ratio of the acceleration coefficient $m$
  %to the drag coefficient $\eta$
  %is
  known as the momentum relaxation time.
  The momentum relaxation time is 
  known to be small for
  particles with low Reynolds numbers,
  indicating 
  a regime in which the acceleration term can be ignored
  entirely.
  (a) Using Newton's Second Law for
  a small momentum relaxation time, 
  demonstrate that a particle confined to a landscape exerting force
  $F^l(\vec{r}_i)$ subject to a time dependent drive $F^d(t)$
  can be modeled with the equation of motion described in 
  Eq.~\ref{eq:motion}. 
  \label{ex:n2l}

@@

\item {\bf Brownian motion}
  is a temperature and viscosity dependent
  phenomena. \cite{}
  Consider the consequences of a non-zero temperature on the simulation, in which particles will collide with invisible particles making up the suspending fluid.  These collisions are more likely at higher temperature due to the increased kinetic energy of the fluid particles, as described by the Maxwell-Boltzmann distribution \cite{}.  In molecular dynamics simulations temperature effects can be modeled by including randomized forces $f^T$ to emulate the motion of particles undergoing Brownian motion.  A distribution of randomizes forces contributes equally in all directions such that the force averaged over a finite time interval is zero $\langle f^T(t) \rangle = 0$.  Random distributions of contain no correlations over the independent variable (time in this case) $\langle f^T(t) f^T(\tau)\rangle = 2 \eta k_B T \delta(t-\tau)$ where the energy $k_B T$ derived from the Boltzmann constant $k_B$ and temperature $T$ has energy units $E_0$.  \cite{Allen2017}

@@

>>>>

We confine the particles using a position dependent 
potential energy function, called a landscape or substrate.
 %The landscape potential energy $U(\vec{r})$ is static 
 %with fixed minima and maxima
 %that are periodic or quasi-periodic,
 %as described in Sec.~\ref{sec:results}.
% 
The landscape is modulated in the y-direction
with the periodic function 
%% %that used in Sec.~\ref{sec:one}
% where 
 \begin{equation}
   U(y) = U_0 \cos{(N_p \pi y / L)}
     \label{eq:ysubstrate}
\end{equation}
 where $N_p$ are the number of periods,
 % $L$ is the period length,
 and $U_0$ is an adjustable parameter
 to set the depth of the minima
 with simulation units of energy $E_0 = 1$. 
 %We define the substrate strength to be
 %the maximum force the substrate can exert on a particle 
 %
 We plot this function in 
 Fig.~\ref{fig:landscape0}
 for $N_p = 3$. %to illustrate the features of
 In Fig.~\ref{fig:landscape0}(a) we show 
 the $x-y$ plane with a contour plot of $U(y)$ %Eq.~\ref{eq:ysubstrate}
 to illustrate
 the 2D potential energy landscape,
 where 
 the maxima are colored red and the minima colored blue.
 The confining force on a particle $i$
 %arises from a potential function 
 %creating a landscape of potential minima and maxima
 %that modify the force on a particle as a function of position
 is calculated as 
 $\vec{F}^{l}_i(\vec{r}_i) = -\nabla U_l(\vec{r}_i)$.
 In Fig.~\ref{fig:landscape0}(b) we plot the function
 $U(y)$ to illustrate how the magnitude
 $\vec{F}^{l}_i$ is calculated from particle position $y_i$.  
 %in the $y-$direction.

@@

\begin{table}[h!]
\centering
\caption{Simulation units and comparable
  experimental values~\cite{Juniper2015}.}
%{\it might be more appropriate to match Juniper parameters more precisely given our focus...}}
\begin{ruledtabular}
\begin{tabular}{c c c } %p{5cm}}
Quantity & Simulation Unit & Experimental range\\
\hline
length &  $a_0 = 1$ & $ a_0 \sim 1.5 \mu m$\\
energy & $E_0 = 1$ & $ < E_0 <$ \\
%electric potential & $V(r_{ij}) = E_0/r_{ij}$ \\
%energy & $E_0 = q^2{Z^*}^2/4\pi \epsilon \epsilon_0 a_0$ \\
%dimensionless interaction strength & $q$ \\
%effective colloidal charge & $Z^*$ \\
%solvent dielectric constant & $\epsilon \epsilon_0$\\
force & $F_0 = E_0 / a_0$ & $<F_0<$\\
velocity &  $ a_0 / \tau $ &  $v \sim 5 \mu m/s $\\
viscosity coefficient & $\eta = F_0 \tau / a_0 $ & $ <\eta<$\\
%juniper uses water–ethanol mixture
time &  $ \tau = \eta a_0 / F_0 $ & $ < \tau <$\\
period & $\lambda = 1.825 a_0$ & $\lambda = 3.5 \mu m$ \\
potential amplitude & $U_0 = 0.058 E_0$ & $U_0 = 75 k_B T \sim 2 J$ \\
temperature & $T_0 = 0$ \footnote{a. see Ex.~\ref{ex:brownian} to include temperature.} & $T \sim 300K $ \\

@@

A single particle %on this landscape
responds to the applied driving force
by moving across the landscape %maxima
synchronized with 
the period of $F^d(t)$.
%applied driving force.
%along the y-direction
%({\it redundant, just refer back to eq. in Model?})
%and $V_{0y}=0.2$.
%This is illustrated in Fig.~\ref{fig:1}(a) where
%the red (blue) regions show local maxima (minima).
%
The numerical implementation of the landscape 
is calculated with Eq.~\ref{eq:ysubstrate} as 
\begin{equation}
  \label{eq:force}
  F^l_y(y) = -A_{p} \sin{(N_p \pi y / L)} 
\end{equation}
where the force is scaled with parameter $A_{p} = N_p \pi U_0/L$.
In this section, we fix the landscape parameters
to $A_{p} = 0.1$ force units
with $N_p=20$ minima %in the landscape
corresponding to a spatial period $\lambda = 1.825 a_0$.

%In Fig.~\ref{fig:0} 
The driven particle adjusts
its motion to traverse
the periodic landscape with the same rhythm as 
the applied time-dependent force $F^d(t)$.
When $F^d(t) > A_p$, a particle can 
overcome the barrier height of the landscape,
and 
the particle hops between minima in the energy landscape.
%
In Fig.~\ref{fig:0}(a)
we plot $F^d(t)$ %defined in Eq.~\ref{eq:drive}.
as a function of time with
%The time independent
constants $F^{dc}=0.1$, 
$F^{ac}=0.05$ and $f=0.01$ cycles per time unit $\tau$.
The temporal period of the driving force is
$T = 1/f = 100 \tau$.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%RAMP
%To avoid transient oscillations at early times, the dc driving force is initially zero and slowly increased %$F^{dc}$. at a rate of $\Delta F^{dc} = 0.01$ every $\Delta t = 4$ time units  to the constant value $F^{dc}=0.1$. In Fig.~\ref{fig:0}(a) we plot $F^{dc}$ in blue to show the increase in  $F^{dc}$ for the first $40$ time units In orange we show  the combined ac and dc drive $F^d(t)$ when the magnitude of the ac applied force is initially zero.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%since $\sin{(0)}=0$.
We mark the maxima of $F^{ac}(t)$
with the
vertical dashed lines %in panels (a) and (b)
-- i.e. $\sin{(\omega t)} = 1$.
%
%
%with the initial driving force at is minimum value 
%$F^d(0) = F^{dc}$. 
%
In Fig.~\ref{fig:0}(b) 
we show the $y-$position of the particle
as a function of time.
We 
normalize $y$ by $\lambda$ %, the spatial period of the potential
to show  particle motion between substrate minima.  
The initial particle position is $y = 0 $. %1 a_0$,
%where the $F^l$ is negative.
%JUST BECAUSE
%causing the particle to move to
%a landscape minima at the start of the simulation.
As the driving force increases,
%the effects of the landscape become less obvious,
the motion of the particle
synchronizes with the driving period $T$.
We measure 
average particle velocity $\langle{v}_y\rangle$
as the average displacement %$\Delta y = y(t_0+T) - y(t_0)$
over the period of the driving force.
The particle moves
in the positive y-direction
through $\Delta y = 2 \lambda$ over the time period $T$,
with 
the average velocity 
$\langle {v}_y \rangle= 2 \lambda f$. % where $T = 1/f$.
The horizontal dashed lines
coincide with %the condition the is
minimum substrate force  $F^l = -A_p$,  
i.e. $\sin{(\pi y / \lambda)} = 1$.
%or $\pi y / \lambda = m \pi/2$, where $m = 1, 5, 7$, {\it etc.}
%Once the particle motion
%coincides with the applied drive,
%indicating synchronization,
The particle synchronizes its motion such that the 
driving force is maximum when the landscape 
force is minimum,
as indicated with %the coincidence of 
with the intersection of
horizontal and vertical lines
and the particle position $y$.
The slope of $y/\lambda$
is proportional to the net force on the particle.
This can be seen in 
Fig.~\ref{fig:0} %we demonstrate this relationship,
where local extrema appear in $y/\lambda$
when $F_{D}(t) \approx 0 $.

@@

%the average velocity of the particle
%is constant 
%despite the overall magnitude
%of $F^{dc}$.
Each step represents a different pattern of hops
between substrate minima
performed by the particle
due to the landscape confinement.  
In Fig.~\ref{fig:1} at %this occurs at 
low $F^{dc}$ the average velocity $\langle v_y \rangle$ is zero.
Since 
$A_p$ is large compared to the extrema of $F^{d}(t)$,
the particle oscillates back and forth
in a single minima with no net velocity.
At higher $F^{dc}$ the particle velocity 
$\langle v_{y} \rangle$ increases in steps of uniform height,
$\langle v_{y} \rangle = n \lambda f$,
where $n$ is an integer.
The step width is non-linear 
and can have a variety of
interesting patterns
such as a devil's staircase related to chaotic dynamics \cite{Bak1986}.
%$\lambda = L/N_p = 36.5/20 = 1.825$ is the spatial period, or wavelength
%of the landscape,
%and $f = 0.01$ cycles per time unit.

@@

In the following section
we include exercises to explore our numerical model,
including the 
to linear drag equation in Ex.~\ref{ex:reynolds}
the equation of motion in Ex.~\ref{ex:n2l},
and 
numerical integration techniques in Ex.~\ref{ex:euler}.
We explore changes in the particle motion
with parameter changes in Ex.~\ref{ex:parameters}.
Methods to visualize the dynamics of the particle motion
are in Ex.~\ref{ex:animation}.
An extended model that includes finite temperature effects
is in Ex.~\ref{ex:brownian}.

@@

@@

  This model should be familiar to readers
  of a standard classical mechanics text, such as Ref.~\cite{Taylor2005}
  when studying 
  models Millikan's oil drops.

  [{\it could certainly do more with this question to make it more computational.  For instance, plot $R$ as a function of any of the parameters $v$, $D$ and $\rho$, include a temperature dependent $\eta(T)$ in the Brownian system.  Suggest students model something big to imagine their own life at low Reynolds number... fairly off topic...}]

@@

\item{\bf Equation of Motion}
  \label{ex:n2l}

  Newton's second law states that
  the acceleration of a particle $i$
  is proportional to 
  the sum of forces on the particle 
  \begin{equation}
  m_i \vec{a}_i = \sum \vec{F}_i
  \label{eq:n2l}
  \end{equation}
  where the constant of proportionality is the
  inertial mass $m_i$
  The addition of a dissipative force in a dynamical model
  on a colloid is typically modeled
  with a drag force proportional to the particle's velocity
  in the opposite direction of motion 
  $\vec{F}^{drag} = - \eta \vec{v}_i$
  where $\eta$ is the drag coefficient.
  The ratio of $m/\eta$ is %has units of time, % and is 
  known as the momentum relaxation time, %.
  and is small for
  particles with low Reynolds numbers.
  When $m/\eta$ is small,
  particle acceleration can be ignored
  entirely.
  (a) Using Newton's Second Law for
  a small momentum relaxation time, 
  demonstrate that a particle confined to a landscape exerting force
  $F^l(\vec{r}_i)$ subject to a time dependent drive $F^d(t)$
  can be modeled with the equation of motion described in 
  Eq.~\ref{eq:motion}.

  {\it this isn't very challenging, but I've see undergraduate students have a ``wow'' factor when they work this out for the first time.}

@@

\item {\bf Brownian motion}
  \label{ex:brownian}
  is a temperature and viscosity dependent
  phenomena. \cite{}
  Consider the consequences of a non-zero temperature on the simulation, in which particles will collide with invisible particles making up the suspending fluid.  These collisions are more likely at higher temperature due to the increased kinetic energy of the fluid particles, as described by the Maxwell-Boltzmann distribution \cite{}.  In molecular dynamics simulations temperature effects can be modeled by including randomized forces $f^T$ to emulate the motion of particles undergoing Brownian motion.  A distribution of randomizes forces contributes equally in all directions such that the force averaged over a finite time interval is zero $\langle f^T(t) \rangle = 0$.  Random distributions of contain no correlations over the independent variable (time in this case) $\langle f^T(t) f^T(\tau)\rangle = 2 \eta k_B T \delta(t-\tau)$ where the energy $k_B T$ derived from the Boltzmann constant $k_B$ and temperature $T$ has energy units $E_0$.  \cite{Allen2017}

@@

\item {\bf Visualizing motion + phase plots}
    \label{ex:animation}
